Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-11 paper/paper.tex
Lots of extra stuff about encoding the logical relation

% \scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
% sentence elsewhere? I'd like to use the previous sentence as a conclusion for
% this section.}
% Finally, due to technical
% details, our mechanized proof about the existence of $\beta\eta$ form does not
% immediately induce a decision procedure for type conversion in
% \lang{}. We discuss this issue and its workaround in
% Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[H]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ |l |  c  | c | c| }
    \hline
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    Syntactic typing         &  83 & =   & = \\
    Untyped reduction        & 344 & =   & = \\
    Neutral and normal forms & -   & 273 & - \\
    Logical relation         & 338 & 430 & - \\
    Semantic soundness       & 192 & 211 & - \\
    Syntactic soundness      & -   & -   & 629 \\
    \hline
    Total                    & 957 & 1341 & 1056 \\
    \hline
  \end{tabular}
  \end{minipage}
  \caption{Nonblank, noncomment lines of code of the Coq Development. The marker = indicates that the line count is the same as the column to the
left. The marker - indicates the file does not
contribute to the total.}
  \label{fig:linecount}
\end{figure}

@@

\subsection{Encoding the logical relation in Coq}
We next discuss specific details of the Coq encoding of the logical relation
presented in Section~\ref{sec:logreldep}.

% First, this definition requires the use of Coq's impredicatve \texttt{Prop}
% sort. In particular, the function $[[F]]$ in \rref{I-Pi} requires the use of
% impredicativity because $[[F]]$ must later be instantiated into something
% defined in terms of the logical relation itself (e.g. in
% Lemma~\ref{lemma:piintroalt}).
In the Coq mechanized proof, % we encode $[[PowerSet STm]]$ as the type \texttt{tm -> Prop}, a predicate over lambda terms. 
the definition of $[[Interp I i A S]]$ has type
\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.

@@

However, if desired, we could consistently replace the use of \texttt{Prop}
with Coq's predicative sort \texttt{Type} in the definition of
$[[Interp I i A S]]$. This alternative definition could be part of
the interpretation for any \emph{finite} number of universes. The use of
\texttt{Type} becomes troublesome only when we attempt to define
$[[InterpR i A S]]$, the top-level logical relation 
(Definition~\ref{fig:logrelrec}) that recursively calls itself at smaller universe
levels. Therefore, the one feature of \lang{} that truly requires
impredicativity is its countable universe hierarchy.

The definition of $[[Interp I i A S]]$ has an almost one-to-one
correspondence to the Coq definition. The main difference is the
specification of $[[I]]$. In
Section~\ref{sec:logreldep}, we define $[[I]]$ as a function over
numbers less than $[[i]]$, the universe level. In Coq, we only require
$[[I]]$ to be a function with the set of natural numbers as its domain.
In the Coq encoding of $[[InterpR i A S]]$, we define $[[I]] \in [[SNat -> PowerSet STm]]$ as follows.
\begin{equation*}
  \begin{split}
    [[I j]] &=
     \begin{cases}
      \ [[{A | exists S , InterpR j A S}]] & \text{when } j < i \\
      \ [[emptyset]] & \text{otherwise}
    \end{cases}
  \end{split}
\end{equation*}
Since $[[I]]$ is only applied to numbers strictly less than $[[i]]$ in
\rref{I-Set}, we can retroactively show that the set we return in the $j
\geq i$ case is junk data that does not affect the result of the logical
relation. This property allows us to recover the simple equation for $[[InterpR i A S]]$ shown in Definition~\ref{fig:logrelrec}.

\Rref{I-PiCoq} shows how \rref{I-Pi} is actually encoded in our
mechanized proof.
\begin{center}
  \drule[width=5in]{I-PiCoq}
\end{center}
Compared to \rref{I-Pi}, \rref{I-PiCoq} replaces the function
$[[F]]$ with a total relation $[[R]]$. The equivalence of these two
rules follows from the fact that the logical relation is a
partial function (Lemma~\ref{lemma:logreldeter}). In set-theoretic
notation, \rref{I-Pi} is more readable. However, if we want to encode
the same rule in Coq, we must encode $[[F]]$ as a relation (with type
\texttt{tm -> (tm -> Prop) -> Prop}) that satisfies the functionality constraint:
\mintinline{coq}{forall a S0 S1, F a S0 -> F a S1 -> S0 = S1}.
In comparison, \rref{I-PiCoq} does not require this side condition and
results in a simpler definition.

We note that we cannot ascribe $[[F]]$ the type \texttt{tm -> (tm ->
  Prop)} since Coq requires functions of such type to be
computable. While defining $[[F]]$ as a computable Coq function rather than a
functional relation does result in a concise encoding of \rref{I-Pi},
we will have trouble instantiating $[[F]]$ with the logical relation,
which is defined as a relation that we prove to be functional, rather
than a computable function.

@@

\subsection{Automation}
\label{sec:automation}
Our Coq mechanization heavily uses automation, supported by the tools
Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.

>>>>

% \scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
% sentence elsewhere? I'd like to use the previous sentence as a conclusion for
% this section.}
% Finally, due to technical
% details, our mechanized proof about the existence of $\beta\eta$ form does not
% immediately induce a decision procedure for type conversion in
% \lang{}. We discuss this issue and its workaround in
% Section~\ref{sec:conversionalgo}.

\section{Mechanization}
\label{sec:logrelmech}

\begin{figure}[h]
  \begin{tabular}{ |l |  c  | c | c| }
    \hline
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    Syntactic typing         &  83 & =   & = \\
    Untyped reduction        & 344 & =   & = \\
    Neutral and normal forms & -   & 273 & - \\
    Logical relation         & 338 & 430 & - \\
    Semantic soundness       & 192 & 211 & - \\
    Syntactic soundness      & -   & -   & 629 \\
    \hline
    Total                    & 957 & 1341 & 1056 \\
    \hline
  \end{tabular}
  \caption{Nonblank, noncomment lines of code of the Coq Development. The marker = indicates that the line count is the same as the column to the
left. The marker - indicates the file does not
contribute to the total.}
  \label{fig:linecount}
\end{figure}

@@

\subsection{Encoding the logical relation}

As discussed, the logical relations $[[Interp I i A S]]$ in
Figures~\ref{fig:logrel} and \ref{fig:logrelopen} are encoded as inductive
definitions indexed by sets of terms representing the semantic inhabitants of
$[[A]]$. In the Coq mechanization, we represent sets of terms as predicates
over terms asserting whether the term is in the set, so $[[S]]$ has type
\texttt{tm -> Prop}, and $[[I]]$ has type \texttt{forall j, j < i -> tm -> Prop},
representing the set of types in universes strictly smaller than $[[i]]$.
Because the overall inductive definition represents a functional relation
between a type at a universe level and its interpretation as a set of terms,
it too lives in \texttt{Prop}.

Due to the limitations of eliminating inductives in \texttt{Prop}, the encoding
of the rule for function types differs from the presentation in \rref{I-Pi},
reproduced below.
%
\begin{mathpar}
  \drule[width=\textwidth]{I-Pi}
\end{mathpar}

While the first and last premises are propositions, $[[S]]$ and $[[F]]$
represent sets of terms, which are not propositions. While this is not an issue
when proving properties of the logical relation, which never require projecting
out these sets, it is an issue when proving the fundamental theorem, and in
particular proving the function case in Lemma~\ref{lemma:stpi}. By the definition
of semantic typing, the induction hypothesis (after some instantiation) gives
%
$$\forall [[a in S]], \exists [[i]], [[S0]] \text{ such that }
  [[InterpR i B {a / x} S0]] \text{ and } [[b a in S0]].$$

We would like to define $[[F]]$ as the function that projects out $[[S1]]$ from
the induction hypothesis, but the existential quantification being a proposition
prevents us from doing so without the axiom of choice. To avoid relying on
classical reasoning, we instead encode $[[F]]$ as a functional relation
$[[R in S * PowerSet STm]]$ such that
$\forall [[a]], \exists [[S0]], [[( a , S0 ) in R]]$ holds. The actual encoding
is given as \rref{I-PiCoq} below.
%
\begin{mathpar}
  \drule[width=\textwidth]{I-PiCoq}
\end{mathpar}

From this rule, we are still able to derive \rref{I-PiAlt} as before. The
benefit of working with the derived rule is that it does not involve $[[R]]$,
which is uniquely determined to be
$[[{ (a , S0 ) | a in S implies Interp I i B { a /x } S0 }]]$ as before.

An alternative to these various encodings in Coq is to use
\emph{induction--recursion} (IR)~\citep{IR}, which is available in Agda~\citep{agda}.
IR permits mutually defining an inductive definition along with a recursive function
on the inductive. We can change our perspective on the logical relation and
view it as two separate pieces: an inductive interpretation $\llbracket A \rrbracket^{i}$
of types, and a recursive interpretation $[[a in InterpR i A]]$ of types as sets of terms.
The rule for function types and its interpretation would be defined as follows.
%
\begin{mathpar}
  \drule[width=\textwidth]{IIR-Pi} \\
  [[b in InterpIR i Pi x : A . B]] \coloneqq
    [[forall a, (# a in InterpIR i A implies b a in InterpIR i B {a / x} #)]]
\end{mathpar}

Because the inductive definition no longer needs to quantify over all sets of terms,
we can work entirely within Agda's lowest \texttt{Set} universe without worrying
about impredicativity or universe polymorphism. However, to define the above,
we still need an intermediate definition involving the interpretation $[[I]]$
of types at $[[j < i]]$ and tie the knot as before. While the same properties
are provable, the inductive--recursive logical relation changes the structure
and the order of these proofs; for instance, a weaker form of functionality
is provable before and is required to prove forward preservation.

Further details of the IR encoding can be found in our Agda mechanization of
the consistency of \lang. Due to the lack of automation support, proofs about
syntactic properties of terms and proofs using well-founded induction on levels
are written manually. Even so, the total number of nonblack, noncomment lines
of code remains small at below 1200 lines, and functional extensionality is
still the only axiom used.

% First, this definition requires the use of Coq's impredicatve \texttt{Prop}
% sort. In particular, the function $[[F]]$ in \rref{I-Pi} requires the use of
% impredicativity because $[[F]]$ must later be instantiated into something
% defined in terms of the logical relation itself (e.g. in
% Lemma~\ref{lemma:piintroalt}).
% In the Coq mechanized proof,
% we encode $[[PowerSet STm]]$ as the type \texttt{tm -> Prop}, a predicate over lambda terms. 
% the definition of $[[Interp I i A S]]$ has type
% \texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.

@@

% However, if desired, we could consistently replace the use of \texttt{Prop}
% with Coq's predicative sort \texttt{Type} in the definition of
% $[[Interp I i A S]]$. This alternative definition could be part of
% the interpretation for any \emph{finite} number of universes. The use of
% \texttt{Type} becomes troublesome only when we attempt to define
% $[[InterpR i A S]]$, the top-level logical relation 
% (Definition~\ref{fig:logrelrec}) that recursively calls itself at smaller universe
% levels. Therefore, the one feature of \lang{} that truly requires
% impredicativity is its countable universe hierarchy.

% The definition of $[[Interp I i A S]]$ has an almost one-to-one
% correspondence to the Coq definition. The main difference is the
% specification of $[[I]]$. In
% Section~\ref{sec:logreldep}, we define $[[I]]$ as a function over
% numbers less than $[[i]]$, the universe level. In Coq, we only require
% $[[I]]$ to be a function with the set of natural numbers as its domain.
% In the Coq encoding of $[[InterpR i A S]]$, we define $[[I]] \in [[SNat -> PowerSet STm]]$ as follows.
% \begin{equation*}
%   \begin{split}
%     [[I j]] &=
%      \begin{cases}
%       \ [[{A | exists S , InterpR j A S}]] & \text{when } j < i \\
%       \ [[emptyset]] & \text{otherwise}
%     \end{cases}
%   \end{split}
% \end{equation*}
% Since $[[I]]$ is only applied to numbers strictly less than $[[i]]$ in
% \rref{I-Set}, we can retroactively show that the set we return in the $j
% \geq i$ case is junk data that does not affect the result of the logical
% relation. This property allows us to recover the simple equation for $[[InterpR i A S]]$ shown in Definition~\ref{fig:logrelrec}.

% \Rref{I-PiCoq} shows how \rref{I-Pi} is actually encoded in our
% mechanized proof.
% \begin{center}
%   \drule[width=5in]{I-PiCoq}
% \end{center}
% Compared to \rref{I-Pi}, \rref{I-PiCoq} replaces the function
% $[[F]]$ with a total relation $[[R]]$. The equivalence of these two
% rules follows from the fact that the logical relation is a
% partial function (Lemma~\ref{lemma:logreldeter}). In set-theoretic
% notation, \rref{I-Pi} is more readable. However, if we want to encode
% the same rule in Coq, we must encode $[[F]]$ as a relation (with type
% \texttt{tm -> (tm -> Prop) -> Prop}) that satisfies the functionality constraint:
% \mintinline{coq}{forall a S0 S1, F a S0 -> F a S1 -> S0 = S1}.
% In comparison, \rref{I-PiCoq} does not require this side condition and
% results in a simpler definition.

% We note that we cannot ascribe $[[F]]$ the type \texttt{tm -> (tm ->
%   Prop)} since Coq requires functions of such type to be
% computable. While defining $[[F]]$ as a computable Coq function rather than a
% functional relation does result in a concise encoding of \rref{I-Pi},
% we will have trouble instantiating $[[F]]$ with the logical relation,
% which is defined as a relation that we prove to be functional, rather
% than a computable function.

@@

\subsection{Automation in Coq}
\label{sec:automation}

Our Coq mechanization heavily uses automation, supported by the tools
Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.
